Iterative Monomorphisation

### 2025-02-13 iterative_monomorphisation.tex
tune
tuning

\usepackage{soul}
\usepackage{multicol}
\usepackage[dvipsnames]{xcolor}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{multirow}
\usepackage[noend]{algorithm2e}
\usepackage{amsmath}
\usepackage{enumitem}

@@

\author{Tanguy Bozec\inst{1,2} \and
Jasmin Blanchette\inst{2}\orcidID{0000-0002-8367-0936}}

\institute{ENS Paris-Saclay, Université Paris-Saclay, France \and
Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}

@@

\begin{abstract}
Monomorphisation can be used to extend some monomorphic provers to support polymorphic logics. We propose an iterative approach, which is necessarily incomplete but which works well in practice. It is implemented in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend. Our evaluation demonstrates that this approach increases Zipperposition's success rate. Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.

\keywords{Polymorphism  \and Monomorphism    \and Automatic theorem proving}
\end{abstract}

@@

One of the main applications of automatic theorem provers is to provide automation to users of proof assistants. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009}, and
Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed only at the top level of formulae. On the other hand, many automatic provers operate only on monomorphic logics. One approach to close this gap is to extend provers to natively support polymorphism, as has been done for Vampire~\cite{bhayat-reger-2020}. This, however, entails a lot of work that needs to be redone for every prover.

The alternative is to translate polymorphic problems to monomorphic problems.
One approach is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic;
we refer to Blanchette et al.~\cite[Section~9]{mono-trans} for an overview of this approach.
Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}. This method relies on heuristically instantiating the formulae's type variables with concrete types.

@@

\SetKwFunction{IterMono}{iterative\_monomorphisation}
\begin{figure}[b]
\begin{quote}
\begin{algorithm}[H]
\Fn(){\(\IterMono(\Phi)\)}{
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{To}{to}

@@

   \Return \(\Mangle(\GenFormulae(\Phi, S))\)
}
\end{algorithm}
\end{quote}
\caption{Pseudocode for iterative monomorphisation}
\label{iter_mono}
\end{figure}

@@

\begin{figure}
\begin{quote}
\begin{algorithm}[H]

@@

\end{algorithm}
\end{quote}
\caption{Pseudocode for formula monomorphisation step}
\label{mono_step}
\end{figure}

@@

\begin{figure}
\begin{quote}
\begin{algorithm}[H]
\SetKwFunction{SubstGen}{matches}
\Fn(){\(\SubstGen(M, N(\varphi))\)}{
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
   \SetKwFunction{Match}{match}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
   \SetKw{Break}{break}

@@

}
\end{algorithm}
\end{quote}
\caption{Pseudocode for match generation}
\label{subst_gen}
\end{figure}

\begin{figure}
\begin{quote}
\begin{algorithm}[H]
\Fn(){\(\GenFormulae(\Phi, S)\)}{
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{To}{to}

@@

   \Return \NewFormulae
}
\end{algorithm}
\end{quote}
\caption{Pseudocode for monomorphic formula generation}
\label{gen_formulae}
\end{figure}

@@

\end{algorithm}
\end{quote}
\caption{Pseudocode for monomorphising substitution generation}
\label{mono_substs}
\end{figure}

>>>>

\usepackage{soul}
\usepackage{multicol}
\usepackage[dvipsnames]{xcolor}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{multirow}
\usepackage[noend]{algorithm2e}
\usepackage{amsmath}
\usepackage{enumitem}
\usepackage{hyperref}

@@

\def\orcid#1{{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}

\author{Tanguy Bozec\inst{1,2}\orcid{TODO} \and
Jasmin Blanchette\inst{2}\orcid{0000-0002-8367-0936}}

\authorrunning{T. Bozec and J. Blanchette}

\institute{ENS Paris-Saclay, Université Paris-Saclay, France \and
%Institute of Informatics,
Ludwig-Maximilians-Universität München, Germany}

@@

\begin{sloppy}
\begin{abstract}
Monomorphisation can be used to extend monomorphic provers to support polymorphic logics. We propose an iterative approach, which is necessarily incomplete but which works well in practice. We implemented it in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend. Our evaluation demonstrates that this approach increases Zipperposition's success rate. Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.

\keywords{Polymorphism  \and monomorphism \and first-order logic
  \and higher-order logic}
\end{abstract}
\end{sloppy}

@@

One of the main applications of automatic theorem provers is to provide automation to users of proof assistants. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009}, and
Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae. On the other hand, many automatic provers operate only on monomorphic logics. One approach to close this gap is to extend provers to natively support polymorphism, as has been done for Vampire~\cite{bhayat-reger-2020}. This, however, entails a lot of work that needs to be redone for every prover.

The alternative is to translate polymorphic problems to monomorphic problems.
One approach to achieve this is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic;
we refer to Blanchette et al.~\cite[Section~9]{mono-trans} for an overview of this approach.
Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}. This method relies on heuristically instantiating the formulae's type variables with concrete types.

@@

\SetKwFunction{IterMono}{iterative\_monomorphisation}
\begin{figure}[t]
%\begin{quote}
\begin{algorithm}[H]
\Fn(){\(\IterMono(\Phi)\)}{
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{To}{to}

@@

   \Return \(\Mangle(\GenFormulae(\Phi, S))\)
}
\end{algorithm}
%\end{quote}
\caption{Pseudocode for iterative monomorphisation}
\label{iter_mono}
\end{figure}

@@

\begin{figure}[t]
%\begin{quote}
\begin{algorithm}[H]

@@

\end{algorithm}
%\end{quote}
\caption{Pseudocode for formula monomorphisation step}
\label{mono_step}
\end{figure}

@@

\begin{figure}[t]
%\begin{quote}
\begin{algorithm}[H]
\SetKwFunction{SubstGen}{matches}
\Fn(){\(\SubstGen(M, N(\varphi))\)}{
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
   \SetKwFunction{Match}{match}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
   \SetKw{Break}{break}

@@

}
\end{algorithm}
%\end{quote}
\caption{Pseudocode for match generation}
\label{subst_gen}
\end{figure}

\begin{figure}[t]
%\begin{quote}
\begin{algorithm}[H]
\Fn(){\(\GenFormulae(\Phi, S)\)}{
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{To}{to}

@@

   \Return \NewFormulae
}
\end{algorithm}
%\end{quote}
\caption{Pseudocode for monomorphic formula generation}
\label{gen_formulae}
\end{figure}

@@

\end{algorithm}
%\end{quote}
\caption{Pseudocode for monomorphising substitution generation}
\label{mono_substs}
\end{figure}
